---
title: "reticulate"
format: html
editor: visual
---

## Python installation
 
For data science, the Anaconda distribution of python is recommended. It includes:

  - Common packages for data science. 
  - Easy management of virtual environments with the package Conda.
  

## Creating a virtual environment named glmhmm

In terminal:

conda create -n glmhmm python=3.9


## Moving to a virtual environment 

In terminal: 

conda activate glmhmm


## Installing packages in an environment

### For packages that are included in conda repository

Once you are in an environment, type in terminal:

conda install numpy 


### For local packages 

https://github.com/zashwood/ssm

In terminal:

cd ssm
pip install numpy cython
pip install -e .

## Moving to a different virtual environment 

In terminal: 

conda deactivate 
conda activate anotherenvironment


## Activating an environment in r

```{r}
library(reticulate)

use_condaenv("glmhmm")
```


## Using python 

```{python}
for i in range(3):
  print(i)
```

## Import a python library

```{python}
import numpy as np
```

```{python}
np.array([1, 2, 3])
```


## Calling python avariable within r

```{python}
a = np.array([1, 2])
```


```{r}
py$a
```


## Calling r variable within python 

```{r}
x <- c(1, 3)
```

```{python}
r.x
```

## You can create a python file and source it

```{r}
source_python("python_code.py")
```

## Python "console
```{r}
repl_python()
```



