---
title: "R Notebook"
output: html_notebook
---

## Exercise 1

The file "dat_contrast.csv" in the folder "data" contains the results of a participant in a detection experiment. On each trial, the participant needs to say "yes" if she can see the stimulus and "no" if she cannot see the stimulus. 

Plot the proportion of detections as a function of contrast and fit a logistic psychometric function.

Which is the estimated value of the slope?

Looking at the plot, for which contrast the proportion of detection is about 50%. 


```{r}
library(tidyverse)
library(broom)

dat_contrast <- read_csv("data/dat_contrast.csv")
```
```{r}
dat_contrast <- dat_contrast |> 
  mutate(response = if_else(response == "yes", 1, 0))
```



```{r}
proportion_contrast <- dat_contrast |> 
  group_by(contrast) |> 
  summarise(prop = mean(response))
```

```{r}
model_contrast <- glm(response ~ contrast, family = binomial(logit), 
                      data = dat_contrast)


```

```{r}
tidy(model_contrast) |> 
  filter(term == "contrast") |> 
  pull(estimate)
```

```{r}
prediction_contrast <- augment(model_contrast, 
                               newdata = tibble(contrast = seq(0, 4, .01)),
                               type.predict = "response")
```

```{r}
proportion_contrast |> 
  ggplot(aes(x = contrast, y = prop)) +
  geom_point() +
  geom_line(data = prediction_contrast, 
            aes(x = contrast, y = .fitted))
```

For 2% contrast, the proportion of detection is about 50%. 


## Exercise 2

An experiment measures the reaction time in response to stimuli that can have two sizes (small and large) and different levels of contrast (the file "dat_rt.csv" is located in the folder data). 

Plot the regression line for the reaction time as a function of contrast for each size considering that the slope is the same for both sizes or that that slopes are different. 


Is there evidence that the slopes are different for the two sizes. 


```{r}
dat_rt <- read_csv("data/dat_rt.csv")
```

```{r}
model_rt_same_slope <- lm(rt ~ contrast + size, data = dat_rt)

reg_line_rt_same_slope <- augment(model_rt_same_slope)

dat_rt |> 
  ggplot(aes(contrast, rt, color = size)) +
  geom_point(position = position_dodge(width = 0.02)) +  #add a small jitter for visualization
  geom_line(data = reg_line_rt_same_slope, 
            aes(x = contrast, y = .fitted))
```

```{r}
model_rt_dif_slope <- lm(rt ~ contrast  * size, data = dat_rt)

reg_line_rt_dif_slope <- augment(model_rt_dif_slope)

dat_rt |> 
  ggplot(aes(contrast, rt, color = size)) +
  geom_point(position = position_dodge(width = 0.02)) +  #add a small jitter for visualization
  geom_line(data = reg_line_rt_dif_slope, 
            aes(x = contrast, y = .fitted))
```


```{r}
summary(model_rt_dif_slope)
```

## Exercise 3

This data frame represents fake data collected from 10 subjects on sleep duration (in hours) and mental well-being scores (on a scale of 1-5).

```{r}
sleep_data <- tibble(Subject = 1:10,
                     Sleep_Duration = c(7.2, 8.5, 6.8, 7.5, 9.1, 5.3, 6.4, 7.9, 8.2, 6.1),
                     Mental_Wellbeing = c(3, 4, 2, 3, 5, 1, 2, 4, 4, 2))
```

  -   Use simple linear regression to examine the relationship between sleep duration and mental well-being.
  -     Report the coefficients of the line of best fit.
  -   Interpret the results in terms of the strength and direction of the relationship between sleep duration and mental well-being.
  -   Create a scatter plot of the data, and add the line of best fit to the plot.
  

```{r}
model_sleep <- lm(Mental_Wellbeing ~ Sleep_Duration, 
                  data = sleep_data)

tidy(model_sleep)
```

```{r}
ggplot()+
  geom_point(data = sleep_data, aes(x = Sleep_Duration, y = Mental_Wellbeing)) +
  geom_line(data = augment(model_sleep), 
            aes(x = Sleep_Duration, y = .fitted))
```

## Exercise 4

The file "dat_contrast_drug.csv" in the folder "data" contains the results of a participant in a detection experiment. On each trial, the participant says "yes" if she can see the stimulus and "no" if she cannot see the stimulus. The participant perform a block of trials under placebo and a block of trials under some drug. Fit a psychometric function model for each block in which the intercept is considered different, but the slope is the same. 


```{r}

dat_contrast_drug_placebo <- read_csv("data/dat_contrast_drug_placebo.csv", 
                                      show_col_types = FALSE)

dat_contrast_drug_placebo <- dat_contrast_drug_placebo |> 
  mutate(response = if_else(response == "yes", 1, 0))

proportions_contrast_drug_placebo <- dat_contrast_drug_placebo |> 
  group_by(block, contrast) |> 
  summarise(prop = mean(response), 
            .groups = "keep")


model_contrast_drug_placebo <- glm(response ~ contrast + block, 
                                   family = binomial(logit), 
                                   data = dat_contrast_drug_placebo)

pred_contrast_drug_placebo <- augment(model_contrast_drug_placebo,
                                      newdata = crossing(block = c("drug","placebo"), 
                                                         contrast = seq(0, 4, .01)), 
                                      type.predict = "response")

proportions_contrast_drug_placebo |> 
  ggplot(aes(x = contrast, y = prop, color = block)) +
  geom_point() +
  geom_line(data = pred_contrast_drug_placebo, 
            aes(x = contrast, y = .fitted, color = block))

```



